Using Graphical Models for Novelty Detection on Semantic Room Classification

### 2011-05-08 semantic-nd/article.tex
Cleaning from novelty detection on given variable of a node.

% first the title is needed
\title{Semantic Novelty Detection with Graphical Models}

@@

\begin{abstract}
This papers presents an approach on implementing novelty detection at semantic level
for indoor room classification in Dora. \emph{Graphical models} are used to model
probabilistic knowledge and novelty detection is implemented on top of them.
The novelty threshold is then optimized using an unconditional probability density
model that is trained from unlabelled data.
The concepts are then generalized to measure novelty introduced by each modelled
variable allowing the system to explain why a class is novel.

@@

There has been several efforts in the area of Artificial Intelligence and Robotics in creating robots
that are able to interact with humans and their environments.
One of the existing problems is a reliable high-level localization method that can be deployed into new
and unknown environments.
In this article we focus on the mapping, using \emph{semantic data}, of indoor environments such as houses and offices to room categories
such as kitchen, corridor, office.
And how to perform novelty detection on them: \emph{detect a new room category that was not present in the labelled data}.

Dora\cite{dora} (CogX: Dora) was used as a base system where this novelty detection would be implemented.
As Dora moves through the environment its \emph{conceptual layer} builds a structural and probabilistic representation of the space
instantiated as a \emph{graphical model}.
That model connects the sensed properties together with the variables used to model the world.
And allows to perform queries on the probabilities of aspects of the environment.
For example: where is most likely to find a cereal box\cite{exploiting}.

@@

The use of graphical models to describe distributions of variables has useful properties.
The edges between the variables can be seen as a kind of filter to the properties of the system.
Properties those that we expect to capture by the conceptual knowledge.
At the same time they are a generative models and therefore allow to calculate the probability
on any given subset of variables on the graph allowing the system to work even when some
information is missing.

@@

\subsection{Factor Graphs}
Although the conceptual layer works with \emph{chain graph}, those can be converted into \emph{factor graphs}.
Those are used through this paper as they provide an easier manipulation in terms of factorization.

A \emph{factor graph} is a bipartite graph connecting nodes representing random variables and factors.
Each factor represents a function dependent only on the variables to where its connected.
As so given factor graph $G$ can be seen as description of a distribution function over a set
of variables obtained by the product of all the factors. In order to represent a probability
a factor $Z$ is defined

@@

\begin{equation}
M_G(x) = \frac{1}{Z}\sum_{V \\ x}{\prod{\phi(...)}}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
\begin{itemize}
\item Define novelty detection 
\item Explain approach with a threshold function
\item Need of a normalizing factor for a threshold with dynamic functions
\item Usage of conditional density
\item Assuming of every output is equally likely
\item Using unlabelled data
\item What happens if the conditional and unconditional graphs differ only by the addition of a node?
\item How to implement that in the case a graph is a tree
\item Novelty detection on any variable of the graphical model!
\end{itemize}

@@

Assuming that the unconditional distributions generates all possible outcome with the same probability
we can model it with $\prod{1/|x_i|}$, where $|x_i|$ denotes the state space size of variable $x_i$.
In graphical model terms, this would be equivalent to build a factor graph $U$ with the same variables but
without any factors as illustrated on \autoref{fig:uniform-prob-graph}.

@@

Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example in very large sets of variables $x$, a large dimension space, the conditional
probability would spread over the tails of the functions and $P_G(x)$ would always yield very
small values.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

@@

\subsection{Generalizing to Novelty Detection on a Variable}
Until now, the presented methods describe how to detect data that belongs to a new class of input
by creating generative models for both conditional and unconditional probabilities distributions.

And although it serves as a method for detecting new classes in the input, it gives no information
in why a class is novel.
The graphical model used by Dora gives a lot of semantic and conceptual information that would be
interesting in exploiting to be able to communicate novelty with a user and to perform
active learning of new classes.
Being able to measure which part of the graph is responsible for triggering novelty would be
a desirable property on the system.
In that sense focus as been made on detecting the impact of any given node.

\begin{figure}[h]
\subfloat[Conditional probability distribution $G$]{\includegraphics[width=0.45\textwidth]{figures/simple-conditional-prob-graph.pdf}}
\qquad
\subfloat[Unconditional probability distribution $G'$]{\includegraphics[width=0.45\textwidth]{figures/simple-unconditional-prob-graph.pdf}}

\caption{\label{fig:measure-room-novelty}In this case the only difference between $G$ and $G'$ is the room category variable
         and the associated factors that only exists in the former.}
\end{figure}

Let the conditional and unconditional probabilities distributions be modeled by graphs
$G$ and $G'$.
And the only difference between those be the non-existence of factors connected to a given
variable $A$ as seen on \autoref{fig:measure-room-novelty}.
Let $N_A$ denote the set of neighbour variables of node $A$.

The ratio: $P_G(x)/P_{G'}(x)$ is then given by:

\begin{equation}
\frac{\sum \prod ... }{\sum \prod ...}  \frac{Z_{G'}}{Z_G}
\end{equation}

The notation $N_v$ is now used to refer to the measure $P_G(x)/P_{G'}(x)$, when $G$ and $G'$ differ only in the variable node $v$ and
its associated factor nodes.
And use it as a novelty measure of the variable $v$, as it attempts to measure on how much adding the variable $v$ helps to explain the
given distribution of values.

By assuming the graph is a tree and blabla, and expanding the calculation of $P_G(x)/P_{G'}(x)$ with the sum-product technique, leads to:

\begin{equation}
N_v(x) = \frac{\sum_{s \in States(v)}x / normalizing_{factor1}}{\prod{...} / normalizing_{factor2}}
\end{equation}

blablablaaa....
I haven't done the math here yet.... but it should give us a method for measuring on how much each variable is contributing to the novelty
threshold.
Also due to the nature of the graphical models it should be possible to calculate in $O(N)$ time for all the $N$ variables.
After the whole variables can be sorted by the contributions and presented to the user explaining where the novelty comes from.

@@

>>>>

% first the title is needed
\title{Using Graphical Models for Novelty Detection on Semantic Room Classification}

@@

\begin{abstract}
This papers presents an approach on implementing novelty detection for indoor room classification
using semantic data.
\emph{Graphical models} are used to model probabilistic knowledge and a novelty threshold is
defined in terms of conditional and unconditional probabilities.
The novelty threshold is then optimized using an unconditional probability density
model trained from unlabelled data.

@@

There has been several efforts in the area of Artificial Intelligence and Robotics in creating
robots that are able to interact with humans and their environments.
One of the existing problems is a reliable high-level localization method that can be deployed
into new and unknown environments.
This article focus on the mapping, using \emph{semantic data}, of indoor environments such as
houses and offices to room categories such as kitchen, corridor, office.
And how to perform novelty detection on them: \emph{detect a new room category that was not
present in the labelled data}.

Dora\cite{dora} (CogX: Dora) was used as a base system where to implement the presented novelty
detection system.
As Dora moves through the environment its \emph{conceptual layer} builds a structural and
probabilistic representation of the space instantiated as a \emph{graphical model}.
That model connects the sensed properties together with the variables used to model the world.
And allows to perform queries on the probabilities of aspects of the environment.
For example: where is most likely to find a cereal box\cite{exploiting}.

@@

The use of graphical models to describe distributions of variables has useful properties.
The edges between the variables can be seen as a kind of filter to the properties of the system.
Properties those that are expected to be captured by the conceptual knowledge.
At the same time they are a generative models and therefore allow to calculate the probability
on any given subset of variables on the graph allowing the system to work even when some
information is missing.

@@

\subsection{Factor Graphs}
Although the conceptual layer works with \emph{chain graphs}, those can be converted into \emph{factor graphs}.
Those are used through this paper as they provide an easier manipulation in terms of factorization.

A \emph{factor graph} $G$ is a bipartite graph connecting a set of nodes $V_G$ representing
random variables and factors $\phi_G$.
Each factor represents a function dependent only on the variables to where its connected.
As so given factor graph $G$ can be seen as description of a distribution function over a set
of variables obtained by the product of all the factors. In order to represent a probability
a normalization factor $Z$ needs to be introduced:

@@

\begin{equation}
P_G(x) = \frac{1}{Z}\sum_{V \\ x}{\prod{\phi(...)}}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
\begin{itemize}
\item Define novelty detection 
\item Explain approach with a threshold function
\item Need of a normalizing factor for a threshold with dynamic functions
\item Usage of conditional density
\item Assuming of every output is equally likely
\item Using unlabelled data
\end{itemize}

@@

Assuming that the unconditional distributions generates all possible outcome with the same probability
we can model it with $\prod{1/\# x_i}$, where $\# x_i$ denotes the cardinality of the state space
of variable $x_i$.
In graphical model terms this is represented to a factor graph $U$ with the variables but
without any factors as illustrated on \autoref{fig:uniform-prob-graph}.

@@

Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example the conditional probability would yield very small values on large sets of variables
$x$ than in small sets due to the spreading over the dimensions of the input space.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

@@

@@

On this paper it was presented how to define a stable novelty threshold function on top of
\emph{probabilistic graphical models} instantiated dynamically from sensed semantic data and
conceptual knowledge.

It was also shown that unlabelled data can be used to approximate unconditional probability and
improve a novelty threshold.
